---
title: 'Coronavirus: creating a HTML table'
output:
  html_document:
    df_print: paged
---

# Coronavirus: creating a HTML table

This notebook documents steps in converting coronavirus data into a HTML page with a searchable table.

The UK government's coronavirus API and the Developer's Guide is at https://coronavirus.data.gov.uk/developers-guide - a separate notebook details the process of working with that.

In this notebook we'll just import some data we've already processed to test out the frontend side of things.

## Import the data

Using the [rio package](https://www.rdocumentation.org/packages/rio/versions/0.5.16) which automatically handles XLSX files

```{r import data}
#grabbing the second sheet and putting in a new object called analysisdf
analysisdf <- rio::import("ltla_analysis.xlsx", sheet = 2)
#Show first few rows
head(analysisdf)
```




## Draw a datatable

Now let's create a datatables interface to look at that. To do that we [use the `DT` package](https://rstudio.github.io/DT/):

```{r}
#This line is commented out because DT is already installed
#install.packages("DT")
#This activates DT in the library, although again it isn't really needed
library(DT)
```


```{r create datatable}
#Now use DT to generate a datatable for the data frame
DT::datatable(analysisdf)
```


```{r}
#Extra parameters set a style and a caption - 
#you can style further see https://rstudio.github.io/DT/
#A filter is added on each column too
#And we specify 30 results per page
DT::datatable(analysisdf, style = 'bootstrap', caption = 'Something you need to know', filter = 'top', options = list(pageLength = 30, autoWidth = TRUE))
```


This will preview in the notebook itself, but you can also generate a HTML version of this notebook by clicking the **Knit** (or **Preview**) button above the notebook, and selecting **Knit to HTML**.

## Creating a separate 'index' page notebook

To keep things simple, create a separate notebook containing the code above. This needs to also define the object being used (again) otherwise it will throw an error.

Here's the code for the other notebook:

```{r code for index notebook, echo=FALSE}
#The echo=FALSE above keeps this code invisible in the HTML output
#Create the data object
analysisdf <- rio::import("ltla_analysis.xlsx", sheet = 2)
#Create the datatable, adding a filter and setting other options
#Add a caption if you want
DT::datatable(analysisdf, 
              style = 'bootstrap', 
              caption = '', 
              filter = 'top', 
              options = list(pageLength = 30, autoWidth = TRUE)
              )
```

(Alternatively you could store this code in an R script and just call that).

We name the notebook 'index.Rmd' because it will generate a correspondingly named HTML file called 'index.html'. This means it can be placed on GitHub and treated as a default homepage.

To generate the HTML file you can run this code:

```{r render notebook as html}
rmarkdown::render("index.Rmd", "all")
```

This cannot be included within the same notebook or it will create a loop and throw an error.

Next we could script the posting of the HTML file to GitHub [using git2r](https://rpubs.com/chrimaho/GitHubAutomation).

## Rounding numbers

We want to round the numbers. [Formatting functions can help](https://blog.rstudio.com/2015/06/24/dt-an-r-interface-to-the-datatables-library/)

```{r}
colnames(analysisdf)
```


```{r round numbers}
DT::datatable(analysisdf, 
              style = 'bootstrap', 
              caption = 'Analysis by the BBC England Data Unit', 
              filter = 'top', 
              options = list(pageLength = 30, autoWidth = TRUE)
              ) %>% 
  formatRound('Rate per 100k population, latest week', 1) %>%
  formatRound('Rate per 100k population, previous week', 1)
```

## Sorting and ranking

We want to add a ranking column. The [rank() function can do this](https://towardsdatascience.com/r-rank-vs-order-753cc7665951).

```{r ranking}
analysisdf
#Add a minus to get descending rank
rank(-analysisdf$`Rate per 100k population, latest week`)
```

This could be used for an additional column, and to sort the results - but we don't want England in there messing up the rankings.

```{r add rank to df}
analysisdf$rank <- rank(-analysisdf$`Rate per 100k population, latest week`)
analysisdf
```


