---
title: "R Notebook"
output: html_notebook
---

I'm going to take stop and search data and identify which Birmingham postcode districts (e.g. B1, B75, etc) have the most stop and search procedures.

First, download the data for West Midlands Police (or another force) from https://data.police.uk/data/

Move it into the same folder as this notebook.

Then import into RStudio.

First let's activat the rio package for importing data.

```{r}
library(rio)
```

Import the data using the import function from rio, and store the data in a new variable

```{r}
wmss2020.08 <- rio::import("2020-08-west-midlands-stop-and-search.csv")
```

Look at the first few rows

```{r}
head(wmss2020.08)
```

```{r}
colnames(wmss2020.08)
```


The data has lat-longs. We can use the postcodes.io API to fetch the postcodes for each one.

First we test one URL

```{r}
testurl <- "http://api.postcodes.io/postcodes?lat=52.55327&lon=-1.900049"
```

## Fetching lists of lats and longs

We can create a vector from any column in a dataframe

```{r}
lats <- wmss2020.08$Latitude
lons <- wmss2020.08$Longitude
```

Now we can loop through that vector

```{r}
for (i in lats){ 
  laturl <- paste("http://api.postcodes.io/postcodes?lat=",i,sep="")
  print(laturl)
  }
```


## Switching to postcodes

Now we look at GP addresses with postcodes from http://media.nhschoices.nhs.uk/data/foi/GP.csv

```{r}
#Import with R
edservices <- rio::import("Scrapereatingdisordersservices.csv")
#Alternative approach with base R
edservices <- read.csv("Scrapereatingdisordersservices.csv", stringsAsFactors = FALSE)
```


```{r}
postcodes <- edservices$V1
```

A typical URL to query the API is: http://api.postcodes.io/postcodes/BL26NT

```{r}
storemylats <- c()
storemylons <- c()

for (postcode in postcodes[2:10]){
  requesturl <- paste("http://api.postcodes.io/postcodes/",postcode,sep="")
  print(requesturl)
  postcodejson <- jsonlite::fromJSON(requesturl)
  jsonlon <- postcodejson$result$longitude
  jsonlat <- postcodejson$result$latitude
  storemylats <- c(storemylats,jsonlat)
  storemylons <- c(storemylons,jsonlon)
  print(storemylats)
  }
```

```{r}
edservices.subset <- edservices[2:10,]
edservices.wlatlon <- cbind(edservices.subset, storemylats, storemylons)

```


