---
title: "R Notebook"
output: html_notebook
---

# My first notebook

Some basic stuff in R

```{r}
myname <- "Paul"
```

Let's store a number

```{r}
myage <- 19
```

## Importing data

To import data in R you can use a function called read.csv()

Using read.csv on its own will display the data but not store it.

```{r}
read.csv("Individual_claims_for_16_17.csv")
```

To *store* data that you read in, you need to **assign** it to a **variable** like this:

```{r}
claims1718 <- read.csv("Individual_claims_for_17_18.csv", sep = ";")
```

The variable that we've just created is a **dataframe** object. This has properties like columns (fields) and rows.

We now import the other CSV file into a second variable:

```{r import 1617}
#import data from CSV 
claims1617 <- read.csv("Individual_claims_for_16_17.csv")
```

## Combining the data

Now we have two dataframes containing the data from our two CSV. We can combine them using the rbind().

```{r}
claims16to18 <- rbind(claims1617, claims1718)
```

## Export the data

We can export as a new CSV using write.csv()

```{r}
write.csv(claims16to18, "alltheclaims.csv")
```

## Different types of variable

We've touched on string variables like "Paul" and numeric variables like 19, and dataframe variables.

Vectors are another type of variable used to store a list.

You create it using the c() function like this:

```{r}
listofages <- c(10,20,60,43)
listofnames <- c("Paul","Darren","Hanna","Ingrid")
listofbooleans <- c(TRUE, FALSE, TRUE, TRUE)
```

A data frame is made up of a number of vectors (columns)

```{r}
mydataframe <- data.frame(listofnames, listofages, listofbooleans)
```

You can extract a vector from a dataframe by using the $ to access it like so:

```{r}
expensenames <- claims1617$MP.s.Name
```


```{r}
statuses <- claims1617$Status
```


```{r}
table(statuses)
```


