---
title: "Getting data for mapping"
output: html_notebook
---

This notebook outlines how to fetch data on the latest crimes from a particular postcode. You can download this notebook and open it in your own R project, then run the code to get up to date data for the same crimes. This data is suitable for mapping using tools like Carto.

# Grabbing crimes for one postcode using the police API

The UK Home Office has an API for police forces in England and Wales. One of the *methods* for that API is '[crimes at a particular location](https://data.police.uk/docs/method/crimes-at-location/)'

There are two example requests at that link. The second fetches data for crimes at a particular latitude and longitude. Those details are included in a URL like this: [https://data.police.uk/api/crimes-at-location?date=2017-02&lat=52.629729&lng=-1.131592](https://data.police.uk/api/crimes-at-location?date=2017-02&lat=52.629729&lng=-1.131592)

The example also includes a `date` parameter, but this is optional. If you omit it, you will get results for the latest month available.

```{r}
#Install the jsonlite package which we need to read JSON 
install.packages('jsonlite')
library('jsonlite')
#Use the fromJSON function from that package to grab data from a URL
crimesmp <- fromJSON("https://data.police.uk/api/crimes-at-location?lat=52.4827963678421&lng=-1.88595522141899")
```

That's great for one location - now we need to get a whole bunch of locations.

Now we could find a big list (for example, all accommodation across the UK) and automate the process. Or we could compile a smaller list (for example, university campuses in one city) manually.

## Finding lat-longs for university campuses

Birmingham City University [has a page listing its campuses](http://www.bcu.ac.uk/about-us/maps-and-campuses). Each campus has its own page, and each of *those* has a link to Google Maps, like this: [https://www.google.co.uk/maps/search/birmingham+city+university+perry+barr/@52.521809,-1.897202,15z?hl=en](https://www.google.co.uk/maps/search/birmingham+city+university+perry+barr/@52.521809,-1.897202,15z?hl=en)

Note that the URL has the lat/long in it: `52.521809,-1.897202`.

We can go to each link and copy the lat long to put in a vector (although not all links have a working Google Maps link):

```{r}
bculatlongs <- c("52.521809,-1.897202","52.483434,-1.885421","52.46884827297614,-1.924323999999956","52.426886,-1.935689","52.481151,-1.903726","52.485907,-1.910819")
bcucampuses <- c("Perry Barr","Millennium Point","City South","BCUIC","Margaret Street","School of Jewellery")
bculocations <- data.frame(bcucampuses,bculatlongs)
```

Curiously, one of the links takes us to [this Google Map of *all* campuses](https://www.google.com/maps/d/viewer?ll=52.46884827297614%2C-1.924323999999956&spn=0.009804%2C0.016093&cd=2&hl=en&f=q&geocode=0%2C52.467677%2C-1.927504&msa=0&source=embed&ie=UTF8&mid=1Iql6zfbHmgtNjdkKLrnLIZlMC-o&z=15). We could extract data from that, too. But not for now...



## Forming the URLs for the police API

Again, we could automate this but for now let's just do it manually line by line:

```{r}
#Use the fromJSON function from that package to grab data from a URL
crimespb <- fromJSON("https://data.police.uk/api/crimes-at-location?lat=52.521809&lng=-1.897202")
crimescs <- fromJSON("https://data.police.uk/api/crimes-at-location?lat=52.46884827297614&lng=-1.924323999999956")
crimesbcuic <- fromJSON("https://data.police.uk/api/crimes-at-location?lat=52.426886&lng=-1.935689")
crimesms <- fromJSON("https://data.police.uk/api/crimes-at-location?lat=52.481151&lng=-1.903726")
crimessoj <- fromJSON("https://data.police.uk/api/crimes-at-location?lat=52.485907&lng=-1.910819")
```

## Combining the results into a single dataset

Now it's time to join the results. Note that some of them had 0 results, so won't combine properly. We can use `rbind` to combine the rows of the 3 that did:

```{r}
allcrimes <- rbind(crimesmp,crimesms,crimespb)
#This throws an error because two fields are lists
#So we need to remove them - given that we already have lat and long the description of the location isn't needed.
crimesmp$location <- NULL
crimesmp$outcome_status <- NULL
crimesms$location <- NULL
crimesms$outcome_status <- NULL
crimespb$location <- NULL
crimespb$outcome_status <- NULL
allcrimes <- rbind(crimesmp,crimesms,crimespb)
```

## Exporting as a spreadsheet we can map online

That's it for now. Let's export it to use in Carto or other mapping software:

```{r}
write.csv(allcrimes,"campuscrimes.csv")
```

