---
title: "Med7373-r-workshop1"
author: "Paul Bradshaw"
date: "02/09/2017"
output: html_document
---

# A first project in R

Firstly, let's break the project down into different steps:

1. Fetch data on school finances from an XLSX file online
2. Clean the data to get rid of unneccessary initial rows
3. Simplify by removing all but the key columns
4. Merge with a second dataset on deprivation
5. Perform some summary statistical analysis
6. Test for correlation between deprivation and finances

## Fetch the data 

We've downloaded the [data from Gov.uk](https://www.gov.uk/government/statistics/income-and-expenditure-in-academies-in-england-2014-to-2015). Now to get it into an R data frame:

```{r}
install.packages("readxl")
library(readxl)
#This specifies the sheet name, and that we need to skip 4 lines before the column headers
#we could instead use sheet=6
academies2016 <- read_excel("SFR32_2017_Main_Tables.xlsx", sheet="Raw Data MATs", skip=4)
```

We're interested in 2 columns: number of pupils and total grant funding. First we need to test the data:

```{r}
summary(academies2016$`Total Grant Funding

£000`)
```

Hm. What about the other:

```{r}
summary(academies2016$`Number of Pupils (FTE)`)
```

Let's try to fix the total column. Could it be the name? To test, let's try accessing the column by index (position) rather than name:

```{r}
head(academies2016[,26])
```

That seems to work (we could also [google the error message](https://stackoverflow.com/questions/39041115/fixing-a-multiple-warning-unknown-column)).

What if we put that column in a separate object and then see if `summary()` generates an error when used on that?

```{r}
totalfunding <- academies2016[,26]
summary(totalfunding)
```

How about we try to add that data back into the original table, with a new name, after deleting the original column and it's problematic name?

```{r}
#Delete column 26 from the original data
academies2016 <- academies2016[-26]
#Add a new column to the original data called 'totalgrantfunding', containing the values that we extracted
academies2016$totalgrandfunding <- totalfunding
#Check it doesn't throw an error like before
summary(academies2016$totalgrandfunding)
```

Now that's fixed, let's see what the correlation is between those:

```{r}
cor(academies2016$`Number of Pupils (FTE)`, academies2016$totalgrandfunding)
```

Ah, another problem. This is probably because of the NA rows. Let's create a subset of the data which doesn't have those:

```{r}
#cbind binds two columns together, we put the result in a new data frame
datatocorrelate <- cbind(academies2016$`Number of Pupils (FTE)`,academies2016$totalgrandfunding, academies2016$`School Name`, academies2016$URN )
#Rename the columns - the second part is a vector of two strings (1 name for each column)
colnames(datatocorrelate) <- c("ftepupils","totalfunding")
#Summarise - we still have those NAs
summary(datatocorrelate$totalfunding)
#Now use na.omit to remove rows with NA
#See https://stackoverflow.com/questions/4862178/remove-rows-with-nas-missing-values-in-data-frame
datatocorrelate <- na.omit(datatocorrelate)
#Check summary again - the NAs are gone
summary(datatocorrelate$totalfunding)
#Now we can correlate
cor(datatocorrelate$ftepupils, datatocorrelate$totalfunding)
```

That's a very strong, correlation, not surprisingly.

## Working out the schools with the most income per pupil

We can create a new column which divides each school's income by that school's number of pupils:

```{r}
datatocorrelate$fundingperpupil <- datatocorrelate$totalfunding/datatocorrelate$ftepupils
#Show a summary
summary(datatocorrelate$fundingperpupil)
```

The median funding per pupil is 5.302 - remember that the figures were in £000s, so that's £5,302 - and in fact we could multiply the figure to make that clearer.

```{r}
datatocorrelate$fundingperpupil <- (datatocorrelate$totalfunding/datatocorrelate$ftepupils)*1000
#Show a summary
summary(datatocorrelate$fundingperpupil)
```

The 'poorest' school ('min') has only £423 per pupil, but the maximum value here is 'Inf', or infinite. Let's take a look at those rows by using `subset`:

```{r}
subset(datatocorrelate, datatocorrelate$fundingperpupil == "Inf")
```

Just two schools, and we can see the reason why: the data says they have no pupils, but still receive some funding. Divide any number by zero and you get 'undefined', expressed here as infinity.

We could choose to investigate those schools further, or choose to remove them in order to get a clearer view on the majority of schools. The former takes us out of R. 

To remove those matching rows we just need to adapt the `subset` expression to say *not* equal to, and assign the results to the same data frame:

```{r}
datatocorrelate <- subset(datatocorrelate, datatocorrelate$fundingperpupil != "Inf")
summary(datatocorrelate$fundingperpupil)
```

Now the max shows one school receiving £407,000 per pupil.

Let's sort the data to bring that to the top:

```{r}
#This make columns in the dataframe callable without having to name the dataframe as well:
attach(datatocorrelate)
#The minus here specifies we want to order in descending, from highest to lowest
orderedbyfundingperpupil <- datatocorrelate[order(-fundingperpupil),]
#Now show the first few rows, which should be the largest
head(orderedbyfundingperpupil)
#Now detach the data frame
detach(datatocorrelate)
```

