---
title: "R exercises"
output: html_notebook
---

This notebook takes you through working with expenses data. First, [download the latest MPs' expenses from IPSA](http://www.theipsa.org.uk/mp-costs/other-published-data/) - *make sure you select Individual Claims* to get the granular data.

## Import data

Now to read that CSV file into an object we can work with in R. Note: the CSV must be in the same folder as your R project.

```{r}
expenses1718 <- read.csv("Individual_claims_for_17_18.csv")
```

## Generate a summary

We can summarise the whole dataset like this:

```{r}
summary(expenses1718)
```

Note how numerical columns are summarised differently to text columns. We can get a summary of just one column like this:

```{r}
summary(expenses1718$Category)
```

Or a numerical one:

```{r}
summary(expenses1718$Amount.Claimed)
```

## Join data

[Download another year's data from IPSA](http://www.theipsa.org.uk/mp-costs/other-published-data/) - remember again to select *Individual Claims*.

Bring it into R in the same way:

```{r}
expenses1617 <- read.csv("Individual_claims_for_16_17.csv")
```

Because we want to add new *rows* with this join, we use `rbind` like so:

```{r}
expenses16to18 <- rbind(expenses1617,expenses1718)
```

## Export the results

That might be all we want to do, then export so we can work with it in Excel.

```{r}
write.csv(expenses16to18, "expenses2016to2017.csv")
```

## Create a subset

Now let's filter this to an MP from one constituency:

```{r}
expensesedgbaston <- subset(expenses16to18, expenses16to18$MP.s.Name == "Stephen Kinnock")
```

## Ask a question

You can ask all sorts of questions...

```{r}
totalexpensesclaimed <- sum(expenses16to18$Amount.Claimed)
totalexpensesclaimed
biggestexpenseclaimed <- max(expenses16to18$Amount.Claimed)
biggestexpenseclaimed
```



