---
title: "Analysing donations data"
output: html_notebook
---

# Analysing donations data

You can download data on election spending and donations from the Electoral Commission website at http://search.electoralcommission.org.uk/. 

We are going to do the following: 

* [Loans between the 2017 and 2019 elections](http://search.electoralcommission.org.uk/Search/Loans?currentPage=1&rows=10&sort=StartDate&order=desc&tab=1&open=filter&et=pp&isIrishSourceYes=true&isIrishSourceNo=true&date=Start&from=&to=&quarters=2019Q1234&quarters=2018Q1234&quarters=2017Q34&register=gb&register=ni&register=none&loanStatus=outstanding&loanStatus=ended&optCols=Register&optCols=CampaigningName&optCols=IsIrishSource&optCols=CompanyRegistrationNumber&optCols=Postcode&optCols=RateOfInterestDescription&optCols=AmountRepaid&optCols=AmountConverted&optCols=AmountOutstanding&optCols=EndDate&optCols=DateRepaid&optCols=DateEcLastNotified&optCols=IsReportedPrePoll&optCols=ReportingPeriodName&optCols=IsAggregation)
* [Donations between the 2017 and 2019 elections](http://search.electoralcommission.org.uk/Search/Loans?currentPage=1&rows=10&sort=StartDate&order=desc&tab=1&open=filter&et=pp&isIrishSourceYes=true&isIrishSourceNo=true&date=Start&from=&to=&quarters=2019Q1234&quarters=2018Q1234&quarters=2017Q34&register=gb&register=ni&register=none&loanStatus=outstanding&loanStatus=ended&optCols=Register&optCols=CampaigningName&optCols=IsIrishSource&optCols=CompanyRegistrationNumber&optCols=Postcode&optCols=RateOfInterestDescription&optCols=AmountRepaid&optCols=AmountConverted&optCols=AmountOutstanding&optCols=EndDate&optCols=DateRepaid&optCols=DateEcLastNotified&optCols=IsReportedPrePoll&optCols=ReportingPeriodName&optCols=IsAggregation)
* [Spending on the 2017 election](http://search.electoralcommission.org.uk/English/Search/Spending?currentPage=1&rows=10&sort=DateIncurred&order=desc&tab=1&open=filter&et=pp&et=ppm&et=tp&et=perpar&et=rd&includeOutsideSection75=true&evt=ukparliament&ev=3568&optCols=CampaigningName&optCols=ExpenseCategoryName&optCols=FullAddress&optCols=AmountInEngland&optCols=AmountInScotland&optCols=AmountInWales&optCols=AmountInNorthernIreland&optCols=DateOfClaimForPayment&optCols=DatePaid)

You should download these yourself - each will just be called 'results.csv' so make sure you rename them accordingly. I've called mine 'loans.csv', 'donations.csv' and 'spending.csv'.

## Import the data

Now to import the data in those files as data frames in R:

```{r}
loans <- read.csv("loans.csv")
```

This should now appear in the *Environment* window in the upper right corner of RStudio.

Let's get a look at the first few rows, using `head()`:

```{r}
head(loans)
```

Each column says `<fctr>` underneath the name. That means each column contains 'factor' variables. This is not really what we want - we want it to have characters (text) and number variables.

To stop that from happening, we need to add another **parameter** to the `read.csv` function when we import the data. That parameter is `stringsAsFactors` and it can be used to specify that we *don't* want to treat strings as factors like so: `stringsAsFactors = FALSE`. 

We separate it from the other parameter (the name of the file) with a comma, like so:

```{r}
loans <- read.csv("loans.csv", stringsAsFactors = FALSE)
```

Now we can check again to see if it's worked:

```{r}
head(loans)
```

That's better - the columns now say `<chr>`. But that includes the column of amounts, which we would hope to treat as numbers.

```{r}
typeof(loans$Value)
```

We could try functions like `as.numeric` or `strtoi` to convert it, but those don't work - instead they convert the amounts to `NA`:

```{r}
strtoi(loans$Value)
as.numeric(loans$Value)
```

The problem is that pound sign. If we can strip that out that might solve our problem.

The function `gsub()` will do this. It substitutes one character for another, and needs at least 3 parameters: what character you want to substitute; what you want to substitute for it (including nothing); and where you want to look (the string or vector of strings).

Let's test it:

```{r}
cleaningvalues <- gsub("£", "", loans$Value)
cleaningvalues
```

That's better - but the commas will still cause us problems, so we can replace those too:

```{r}
cleaningvalues <- gsub(",", "", cleaningvalues)
cleaningvalues
```

Those amounts can now be converted using `as.numeric()`:

```{r}
as.numeric(cleaningvalues)
```

Let's assign that back to the main dataset as a new column:

```{r}
loans$valueasnum <- as.numeric(cleaningvalues)
head(loans)
```





## Nesting

By the way, we could also nest each of those steps within each other like so to do them all at once (but don't worry if this looks really complicated!):

```{r}
as.numeric(gsub(",","",(gsub("£", "", loans$Value))))
```

## Importing the other data

We can repeat some of those steps with the other data:

```{r}
donations <- read.csv("donations.csv", stringsAsFactors = FALSE)
spending <- read.csv("spending.csv", stringsAsFactors = FALSE)
```


