---
title: "Analysing the British Election Study"
output: html_notebook
---

# Analysing the British Election Study

The British Election Study contains data on surveys of voters going back a number of years. The panel study is based on over 90,000 interviews and is supplied as a large file in the SPSS or STATA formats. 

R is well suited for dealing with both the format of the data, and its size.

[You can download any of the BES's panel studies as a large zip file here](https://www.britishelectionstudy.com/data-objects/panel-study-data/) - you may have to register for a free account first.

## Importing the data

When you unzip the zip file you will have a file ending in .sav - we can use a package like `rio` which can recognise and deal with that file type - or `haven` which has a specific function for importing it (`read_sav()`).

```{r}
#library(rio)
#bes19 <- rio::import('BES2019_W16_Panel_v0.3.sav')
library(haven)
bes19.haven <- haven::read_sav('BES2019_W16_Panel_v0.3.sav')
```

You will see that not only does it have over 90,000 rows but also thousands of columns, which is why it's so big.

```{r}
colnames(bes19.haven)
```

```{r}
table(bes19.haven[156])
table(bes19.haven[155])
table(bes19.haven$partyContactConW2)
table(bes19.haven$partyContactConW2)

```


```{r}
head(bes19.haven)
```

```{r}
summary(bes19.haven$starttimeW1)
```

