---
title: "Using parlitools"
output: html_notebook
---

# Using parlitools

[The `parlitools` package](https://cran.r-project.org/web/packages/parlitools/vignettes/introduction.html) "is a collection of datasets, maps and data retrieval functions for analysing UK politics".

```{r}
install.packages("parlitools")
?parlitools
library(dplyr)
library(ggplot2)

census_11 <- parlitools::census_11

bes_2017 <- parlitools::bes_2017

elect_results <- left_join(census_11, bes_2017)

degree_plot <- ggplot(elect_results, aes(y=lab_17, x=degree)) +
  geom_point(alpha=0.75) +
  geom_smooth(size=1.75, colour = "#DC241F") +
  ylab("Share of Votes Cast for Labour") + 
  xlab("Percentage of Population with a University Degree")

degree_plot
```

```{r}
install.packages("cartogram")

library(leaflet)
library(sf)
library(htmlwidgets)
library(dplyr)
library(parlitools)
library(cartogram)

west_hex_map <- parlitools::west_hex_map

party_colour <- parlitools::party_colour

elect2017 <- parlitools::bes_2017

elect2017_win_colours <- left_join(elect2017, party_colour, by = c("winner_17" ="party_name")) #Join to current MP data

gb_hex_map <- right_join(west_hex_map, elect2017_win_colours, by = c("gss_code"="ons_const_id")) #Join colours to hexagon map

gb_hex_map <- as(gb_hex_map, "Spatial")

gb_hex_map <- as(gb_hex_map, "SpatialPolygonsDataFrame")
gb_hex_map
gb_hex_map$majority_17 <- round(gb_hex_map$majority_17, 2)

gb_hex_map$turnout_17 <- round(gb_hex_map$turnout_17, 2)

gb_hex_map$marginality <- (100-gb_hex_map$majority_17)^3

gp_hex_scaled <- cartogram_cont(gb_hex_map, 'marginality', itermax = 5)
gp_hex_scaled

# Creating map labels
labels <- paste0(
  "Constituency: ", gp_hex_scaled$constituency_name.y, "</br>",
  "Most Recent Winner: ", gp_hex_scaled$winner_17, "</br>",
  "Most Recent Majority: ", gp_hex_scaled$majority_17, "%","</br>",
  "Turnout: ", gp_hex_scaled$turnout_17, "%"
) %>% lapply(htmltools::HTML)

# Creating the map itself
leaflet(options=leafletOptions(
  dragging = FALSE, zoomControl = FALSE, tap = FALSE,
  minZoom = 6, maxZoom = 6, maxBounds = list(list(2.5,-7.75),list(58.25,50.0)),
  attributionControl = FALSE),
  gp_hex_scaled) %>%
  addPolygons(
    color = "grey",
    weight=0.75,
    opacity = 0.5,
    fillOpacity = 1,
    fillColor = ~party_colour,
    label=labels) %>% 
  htmlwidgets::onRender(
    "function(x, y) {
        var myMap = this;
        myMap._container.style['background'] = '#fff';
    }")%>% 
  mapOptions(zoomToLimits = "first")
```


```{r}
gp_hex_scaled
```


## Reverse engineering the map

The `west_hex_map` is the base object - we can map that by replacing `gp_hex_scaled` in the code that draws the map:

```{r}
# Basic hex map
leaflet(options=leafletOptions(
  dragging = FALSE, zoomControl = FALSE, tap = FALSE,
  minZoom = 6, maxZoom = 6, maxBounds = list(list(2.5,-7.75),list(58.25,50.0)),
  attributionControl = FALSE),
  west_hex_map) %>%
  addPolygons(
    color = "grey",
    weight=0.75,
    opacity = 0.5,
    fillOpacity = 1,
    fillColor = ~party_colour,
    label=labels) %>% 
  htmlwidgets::onRender(
    "function(x, y) {
        var myMap = this;
        myMap._container.style['background'] = '#fff';
    }")%>% 
  mapOptions(zoomToLimits = "first")
```


We can change bits to see what they control. For example the background colour in `myMap._container.style['background'] = '#fff';`:

```{r}
# Basic hex map
leaflet(options=leafletOptions(
  dragging = FALSE, zoomControl = FALSE, tap = FALSE,
  minZoom = 6, maxZoom = 6, maxBounds = list(list(2.5,-7.75),list(58.25,50.0)),
  attributionControl = FALSE),
  west_hex_map) %>%
  addPolygons(
    color = "grey",
    weight=0.75,
    opacity = 0.5,
    fillOpacity = 1,
    fillColor = ~party_colour,
    label=labels) %>% 
  htmlwidgets::onRender(
    "function(x, y) {
        var myMap = this;
        myMap._container.style['background'] = 'red';
    }")%>% 
  mapOptions(zoomToLimits = "first")
```
We can also change `color = "grey"` - this seems to affect the borders between polygons. And `fillColor = ~party_colour,` which affects the colour of the polygons.


```{r}
# Basic hex map

leaflet(options=leafletOptions(
  dragging = FALSE, zoomControl = FALSE, tap = FALSE,
  minZoom = 6, maxZoom = 6, maxBounds = list(list(2.5,-7.75),list(58.25,50.0)),
  attributionControl = FALSE),
  west_hex_map) %>%
  addPolygons(
    #Change this line
    color = "red",
    weight=0.75,
    opacity = 0.5,
    fillOpacity = 1,
    #And this line
    fillColor = 'yellow',
    label=labels) %>% 
  htmlwidgets::onRender(
    "function(x, y) {
        var myMap = this;
        myMap._container.style['background'] = 'white';
    }")%>% 
  mapOptions(zoomToLimits = "first")
```

Can we get the colours in?

```{r}
gb_hex_map <- right_join(west_hex_map, elect2017_win_colours, by = c("gss_code"="ons_const_id")) #Join colours to hexagon map
elect2017_win_colours
```



```{r}
# Basic hex map
leaflet(options=leafletOptions(
  dragging = FALSE, zoomControl = FALSE, tap = FALSE,
  minZoom = 6, maxZoom = 6, maxBounds = list(list(2.5,-7.75),list(58.25,50.0)),
  attributionControl = FALSE),
  gb_hex_map) %>%
  addPolygons(
    #Change this line
    color = "grey",
    weight=0.75,
    opacity = 0.5,
    fillOpacity = 1,
    #And this line
    fillColor = -party_colour,
    label=labels) %>% 
  htmlwidgets::onRender(
    "function(x, y) {
        var myMap = this;
        myMap._container.style['background'] = 'white';
    }")%>% 
  mapOptions(zoomToLimits = "first")
```


```{r}
colnames(gb_hex_map)
gb_hex_map <- left_join(gb_hex_map, party_colour, by = c("winner_17" ="party_name")) #Join to current MP data

```

## Another map

This is from [the documentation](https://www.rdocumentation.org/packages/parlitools/versions/0.2.0/vignettes/introduction.Rmd):

```{r}
library(leaflet)
library(sf)
library(htmlwidgets)
library(dplyr)
library(hansard)
library(mnis)
library(parlitools)

west_hex_map <- parlitools::west_hex_map

party_col <- parlitools::party_colour

mps <- mps_on_date("2017-06-20")

mps_colours <- left_join(mps, party_col, by = "party_id") #Join to current MP data

west_hex_map <- left_join(west_hex_map, mps_colours, by = "gss_code") #Join colours to hexagon map

# Creating map labels
labels <- paste0(
  "<strong>", west_hex_map$constituency_name, "</strong>", "
",
  "Party: ", west_hex_map$party_name, "
",
  "MP: ", west_hex_map$display_as, "
",
  "Most Recent Result: ", west_hex_map$result_of_election, "
",
  "Current Majority: ", west_hex_map$majority, " votes"
) %>% lapply(htmltools::HTML)

# Creating the map itself
leaflet(options=leafletOptions(
  dragging = FALSE, zoomControl = FALSE, tap = FALSE,
  minZoom = 6, maxZoom = 6, maxBounds = list(list(2.5,-7.75),list(58.25,50.0)),
  attributionControl = FALSE),
  west_hex_map) %>%
  addPolygons(
    color = "grey",
    weight=0.75,
    opacity = 0.5,
    fillOpacity = 1,
    fillColor = ~party_colour,
    label=labels)  %>% 
  htmlwidgets::onRender(
    "function(x, y) {
        var myMap = this;
        myMap._container.style['background'] = '#fff';
    }")
```

