---
title: "SQLdf example"
output: html_notebook
---

## Using SQL in R: the `sqldf` package

One way to use SQL queries in R is to install the `sqldf` package (think of it as SQL data frame). Do that first:

```{r}
install.packages("sqldf")
library(sqldf)
```

Next we need some data to run our query on. Here's our familiar expenses data:

```{r}
exes1718 <- read.csv("exes1718.csv")
```

## The first query: a filter

Now for a basic query. To use a SQL query in R you just use the `sqldf` function and put your SQL query inside the brackets that follow. Note that the query is enclosed in quotation marks, so if you want to specify a string in your query, you need to make sure that string uses a different type of quotation mark:  

```{r}
sqldf('select * from exes1718 where Category is "Accommodation"')
```

That query shows the results, but let's store them instead, in a new object:

```{r}
accommclaims <- sqldf('select * from exes1718 where Category is "Accommodation"')
```

Of course the same result can be achieved by using the `subset` function, but you may find this more intuitive. Likewise, removing columns is also more intuitive with SQL...

## Selecting specific columns

The `select *` command in SQL selects all columns, but you can select them each individually instead:

```{r}
dateandcatonly <- sqldf("select Date, Category from exes1718")
```

Note that SQL struggles with some column names where R replaces spaces with full stops. You may have to rename these first:

```{r}
#Show the column names
colnames(exes1718)
#The fourth one seems to have full stops in it. Let's check:
colnames(exes1718)[4]
```


```{r}
#Now rename that column name
colnames(exes1718)[4] <- "mpname"
colnames(exes1718)[5] <- "mpconstituency"
colnames(exes1718)[7] <- "expensetype"
colnames(exes1718)[8] <- "description"
colnames(exes1718)[16] <- "amountclaimed"
colnames(exes1718)[17] <- "amountpaid"
#And check it
colnames(exes1718)
```

## Calculations using SQL: counts

 SQL allows us to perform calcuations on the data too, such as counting matches and calculating totals, averages, and so on...

Here are 2 SQL queries asking to count how many entries in the data frame are above 100, and how many are below 100. 
 
```{r}
sqldf('select count(*) from exes1718 where amountpaid >100')
sqldf('select count(*) from exes1718 where amountpaid <100')
```
How would you adapt that query to ask how many are exactly 100? Or above 1000?

### Calculations: sums

To add them, just use `sum` instead of `count` - but this time you need to make sure you are using `sum` with a numerical column:

```{r}
sqldf('select sum(amountpaid) from exes1718 where amountpaid >100')
```

Note that the column you are testing, and the column you are adding, do *not* have to be the same. We can write another query that adds up the amount paid where the category is accommodation:

```{r}
sqldf('select sum(amountpaid) from exes1718 where category is "Accommodation"')
```

### Calculations: averages

Besides counts and sums, we can also calculate averages using `avg` for the mean:

```{r}
sqldf('select avg(amountpaid) from exes1718 where category is "Accommodation"')
```

Or `median` for the middlemost value:

```{r}
sqldf('select median(amountpaid) from exes1718 where category is "Accommodation"')
```

...and the `mode` (most commonly occurring number):

```{r}
sqldf('select mode(amountpaid) from exes1718 where category is "Accommodation"')
```

### Calculating biggest and smallest amounts

Not surprisingly, `max` and `min` can be used to find those values. Here we switch to a different criteria too:

```{r}
sqldf('select max(amountpaid) from exes1718 where mpname is "Adrian Bailey"')
```

And for minimum:

```{r}
sqldf('select min(amountpaid) from exes1718 where mpname is "Adrian Bailey"')
```


### Statistical queries

We can also ask for things like the standard deviation, an indication of how widely numbers vary:

```{r}
sqldf('select stdev(amountpaid) from exes1718 where description is "Office Costs"')
```

## Grouping results - pivot tables using `group by`

Of course we probably don't want to have to do this for every category or MP. So the `group by` command in SQL is useful to perform a calculation for all categories, places or people with a breakdown for each:


```{r}
sqldf('select sum(amountpaid) from exes1718 group by category')
```

Note that the categories themselves in the above query aren't shown. Why? Because we haven't selected them! Let's rectify that:

```{r}
sqldf('select category, sum(amountpaid) from exes1718 group by category')
```

You can even add other calculations to make a bigger table:

```{r}
sqldf('select category, stdev(amountpaid), max(amountpaid), min(amountpaid), avg(amountpaid),  median(amountpaid) from exes1718 group by category')
```

It is interesting that staffing seems to have the widest variation even though its maximum and minimum numbers are not unusually large or small.

### Sorting results: `order by`

So it looks like staffing costs vary most. But we can make it easier by adding an `order by` command:

```{r}
sqldf('select category, stdev(amountpaid) from exes1718 group by category order by stdev(amountpaid) desc')
```


## String queries: `like`

So far we've matched text columns based on an exact match like `category is "Accommodation"`. But what if we want to grab partial matches where a word is somewhere in that column? The `like` command used with the `%` operator as a wildcard allows us to do that:

```{r}
sqldf('select mpname, mpconstituency from exes1718 where mpconstituency like "%Birmingham%" group by mpname')
```

We can broaden the query further by adding `or` like so:

```{r}
sqldf('select mpname, mpconstituency from exes1718 where mpconstituency like "%Birmingham%" or mpconstituency like "%Sutton Coldfield%" group by mpname')
```

Note that *after* `or` you need to repeat the name of the column and `like` again.

## Saving and exporting results

Of course any results can be stored in an R object and/or exported like so:

```{r}
# How to create a new object
exesstdevbycategory <- sqldf('select category, stdev(amountpaid) from exes1718 group by category')
# How to write results as a CSV
write.csv(sqldf('select category, stdev(amountpaid) from exes1718 group by category'), "exesbycategorystdev.csv")
```

